Manuscript 1 code

### 2024-10-02 manuscript_1_code.Rmd
added functional group work from Lauren up to figure 3b

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(here)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(vegan)
library(glmmTMB)
library(car)
library(effects)
library(DHARMa)
library(stringr)
library(MuMIn)
library(codyn)
library(nlme)
library(mgcv)
library(emmeans)
library(dplyr)
```

@@

### site richness
```{r site richness mod}
site_cover_mod <- glmmTMB(cover_trans ~ richness*habitat + (1|site_habitat) + (1|year), family = beta_family(), data = alpha_diversity_site_macro_with_cover)
summary(site_cover_mod)
Anova(site_cover_mod)
#                    Chisq Df Pr(>Chisq)    
# richness         58.9195  1  1.643e-14 ***
# habitat           6.6247  3    0.08487 .  
# richness:habitat  3.2967  3    0.34811    
hist(residuals(site_cover_mod)) # some positive skew, not the worst thing ever
plot(residuals(site_cover_mod) ~ fitted(site_cover_mod)) # a little heteroscedastic but can't be less than 0
r.squaredGLMM(site_cover_mod) # 0.2212534, 0.5990825
emtrends(site_cover_mod, pairwise ~ habitat, var = "richness") # no difference between the habitats
```

@@

@@

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(here)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(vegan)
library(glmmTMB)
library(car)
library(effects)
library(DHARMa)
library(stringr)
library(MuMIn)
library(codyn)
library(nlme)
library(mgcv)
library(emmeans)
library(dplyr)
library(janitor)
```

@@

@@

## Figure 3b - functional groups

### Setup
```{r read and prep functional group data}
algae_fg_raw2 <- read_csv(here("data","algae_functional_groupings_092724.csv"))

#clean column names
algae_fg <- algae_fg_raw2 %>%
  janitor::clean_names()

#change species names to be lowercase & have underscore 
algae_fg$species <- gsub(" ", "_", tolower(algae_fg$species))
algae_fg$morphology <- trimws(algae_fg$morphology)
algae_fg$morphology <- gsub(" ", "_", algae_fg$morphology)

#rename species column to taxa

algae_fg <- algae_fg %>%
  select(species, morphology) %>%
  dplyr::rename(taxa = species) %>%
   dplyr::rename(functional_group = morphology)

unique(algae_fg$functional_group)

#merge functional groups with 
macro_functional_groups <- merge(macro_long_data, algae_fg, all.x = TRUE)

taxa_sum_check <- macro_functional_groups %>%
  group_by(location, year) %>%
  summarise(sum_macrocover = sum(prop_cover))

macro_functional_groups %>%
  filter(location == "lter_2_backreef_algae_transect_1_quad_1") %>%
  filter(year == "2022")

#summarizing functional groups at the quad level
fg_summary <- macro_functional_groups %>%
  group_by(location, year, functional_group, habitat, site) %>%
  summarise(sum_fg_cover = sum(prop_cover))

max(fg_summary$sum_fg_cover)

#summarizing functional groups at the site level
#NOTE that this is ADDED and not averaged. This data is only used to find richness at the site level. LE did this as adding and as a sum, both end up the same from a richness perspective. 
fg_summary_site_SUM <- macro_functional_groups %>%
  group_by(site, year, habitat, functional_group) %>%
  summarise(sum_fg_cover = sum(prop_cover))

#don't need to use this, was just to check both methods
#fg_summary_site_AVG <- macro_functional_groups %>%
  #group_by(site, year, habitat, functional_group) %>%
  #summarise(mean_fg_cover = mean(prop_cover))

#fg_summary %>%
  #filter(sum_fg_cover==1)

#QAQC
fg_summary_qaqc <- macro_functional_groups %>%
  group_by(location, year, functional_group) %>%
  summarise(sum_fg_cover = sum(prop_cover))

fg_qaqc <- fg_summary_qaqc %>%
  group_by(location, year) %>%
  summarise(sum_fg_check = sum(sum_fg_cover))

check_them <- merge(taxa_sum_check, fg_qaqc)

check_them %>%
  filter(all.equal(sum_macrocover, sum_fg_check) != TRUE)

#pivot wide at quad level
fg_summary_wide <- fg_summary %>%
  pivot_wider(names_from = "functional_group", values_from = "sum_fg_cover")

#richness at every quad/year combination 
fg_summary_wide$fg_richness_quad <- vegan::specnumber(fg_summary_wide[5:14])

unique(fg_summary_wide$fg_richness_quad)
#0 to 4

##### REPEAT AT SITE LEVEL
#pivot wide at site level
fg_summary_site_wide_SUM <- fg_summary_site_SUM %>%
  pivot_wider(names_from = "functional_group", values_from = "sum_fg_cover")

#richness at every site/year combination ---> NOT AVERAGED,
fg_summary_site_wide_SUM$fg_richness_site <- vegan::specnumber(fg_summary_site_wide_SUM[4:13])

fg_summary_site_wide <- fg_summary_site_wide_SUM %>%
   mutate(site_habitat = paste(site, habitat, sep = "_"))
```

### model
```{r functional richness cover mod quad}
#merge Noam's data with functional group summary 
functional_richness_quad_cover <- merge(alpha_diversity_quad_macro_with_cover, fg_summary_wide,
                                        by = c("location", "year", "site", "habitat"))

cover_mod_fg <- glmmTMB(cover_trans ~ fg_richness_quad*habitat + (1|site/location) + (1|year), data = functional_richness_quad_cover)

summary(cover_mod_fg)
car::Anova(cover_mod_fg)
#                              Chisq Df Pr(>Chisq)    
# fg_richness_quad         10323.133  1  < 2.2e-16 ***
# habitat                     87.166  3  < 2.2e-16 ***
# fg_richness_quad:habitat    77.758  3  < 2.2e-16 ***
hist(residuals(cover_mod_fg)) # a little positive skew, not the worst

r.squaredGLMM(cover_mod_fg) # 0.39, 0.51
emtrends(cover_mod_fg, pairwise ~ habitat, var = "fg_richness_quad") 
#all habitats are different from forereef 17

## figure 3b plot setup
```{r figure 3b setup}
diversity_ranges_quad_fg <- functional_richness_quad_cover %>%
  group_by(habitat) %>%
  summarise(min_fg_richness_quad = min(fg_richness_quad),
            max_fg_richness_quad = max(fg_richness_quad),
            min_cover= min(cover_trans),
            max_cover = max(cover_trans))

cover_emm <- emmip(cover_mod_fg,
                        habitat ~ fg_richness_quad,
                        at = list(fg_richness_quad = seq(0,10,1)), plotit = F, CIs = T)

filtered_cover_effects <- cover_emm %>%
  left_join(diversity_ranges_quad_fg, by = c("habitat")) %>%
  filter(fg_richness_quad >= min_fg_richness_quad & fg_richness_quad <= max_fg_richness_quad)

## figure 3b plot

```{r figure 3b plot}
(figure_3b <- 
  ggplot() +
  geom_point(data = functional_richness_quad_cover,
             aes(x = fg_richness_quad, y = cover_trans, colour = habitat), size = 1, alpha = 0.005) + 
  geom_line(data = filtered_cover_effects, aes(x = fg_richness_quad, y = inv_logit(yvar), colour = habitat)) +
  geom_ribbon(data = filtered_cover_effects, aes(x = fg_richness_quad, ymin = inv_logit(LCL), ymax = inv_logit(UCL), fill = habitat),
              alpha = 0.3) +
  scale_colour_manual(values = habitat_colours) +
  scale_fill_manual(values = habitat_colours) +
  labs(y = "", x = "Functional Richness", title = "3b.") +
  theme_bw() + 
  theme(axis.title.x = element_text(size=24), 
        axis.title.y = element_text(size=24), 
        axis.text.x = element_text(size=22), 
        axis.text.y = element_text(size=22),
        legend.text = element_text(size=22),
        plot.title = element_text(size=26), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_rect(fill = "white"),
        legend.title = element_blank(),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

#ggsave(filename = "output/figure_3b_quad_v2.png", figure_3b, height = 10, width = 14)
```

@@

### model at site level
```{r site richness mod}
site_cover_mod <- glmmTMB(cover_trans ~ richness*habitat + (1|site_habitat) + (1|year), family = beta_family(), data = alpha_diversity_site_macro_with_cover)
summary(site_cover_mod)
Anova(site_cover_mod)
#                    Chisq Df Pr(>Chisq)    
# richness         58.9195  1  1.643e-14 ***
# habitat           6.6247  3    0.08487 .  
# richness:habitat  3.2967  3    0.34811    
hist(residuals(site_cover_mod)) # some positive skew, not the worst thing ever
plot(residuals(site_cover_mod) ~ fitted(site_cover_mod)) # a little heteroscedastic but can't be less than 0
r.squaredGLMM(site_cover_mod) # 0.2212534, 0.5990825
emtrends(site_cover_mod, pairwise ~ habitat, var = "richness") # no difference between the habitats
```

@@

## Figure 4b - functional group richness stability
### Setup
```{r functional richness through time at quad level}
fg_summary_wide_meta <- merge(fg_summary_wide, meta_df, by = c("location", "site", "habitat", "year") )
fg_summary_wide_meta_through_time <- fg_summary_wide_meta %>%
  group_by(location, habitat, site) %>%
  summarise(fg_richness_mean = mean(fg_richness_quad),
            fg_richness_se = sd(fg_richness_quad)/n())

#each quad only has ONE value, averaged through time 
#1200 quads in this data set
#this is used in figure 4 things
```
